# Combining Datasets

## About this chapter

1. Questions:
- How do I combine dataframes?
2. Objectives:
- Understanding keys
- Explore `join` functions
3. Keypoints:
- Dataframes get joined on key columns. The rows that are retained depends on the type of `join` performed


## Joining

Often you will want to combine data contained in more than one dataset. In this section we will look at the functions that help you do that.

### Key columns

The joining operation depends on the two datasets having some values in some column in common. The column in each dataset that allows you to combine columns is the key column. Consider these dataframes
```{r, echo=FALSE, warning=FALSE,message=FALSE}
library(dplyr)
```

```{r}
band_members

band_instruments
```

The key column in `band_members` is `name` and in `band_instruments` it is `artist`.

## Join functions

The join functions in `dplyr` all have a common syntax. Here's `left_join()` as an example


```{r, eval = FALSE}
left_join(x, y, by.x = key_column_x, by.y = key_column_y)
```

The `x` is the first (or left) dataframe, the `y` is the second (or right) dataframe and `by.x` and `by.y` are the key columns to use.

### left_join()

The most common sort of join is the left join. This takes one dataframe, considers it to be on the left of the join and combines the second dataframe to it, skipping rows in the right dataframe that have nowhere to join

```{r}
left_join( band_members, band_instruments, by.x = name, by.y = artist )
```

Note how the `band_member` `Keith` goes missing because it isn't in the `x` (left) dataframe. Note also how the key column name has come from the `x` dataframe.

### right_join()

`right_join()` is the complementary function. 

```{r}
right_join( band_members, band_instruments, by.x = name, by.y = artist)
```

See how this time `Keith` is retained as we're joining to the right table, but as he has no entry in the left table, an `NA` is used to fill the missing value.

### inner_join()

`inner_join()` keeps only rows that are completely shared

```{r}
inner_join( band_members, band_instruments, by.x = name, by.y = artist)
```

### full_join()

`full_join()` joins all rows as well as possible, generating `NA` as appropriate.

```{r}
full_join( band_members, band_instruments, by.x = name, by.y = artist)
```

## Binding operations

These allow you to paste dataframes together.

`bind_rows()` sticks them together top-to-bottom

```{r}
bind_rows(band_members, band_members)
```

Note the column names need not be identical for this to work. `NAs` are propogated as required.

```{r}
bind_rows(band_members, band_instruments)
```

`bind_cols()` sticks dataframes together side-by-side/
```{r}
bind_cols(band_members, band_instruments)
```

Note how it doesn't do any sensible matching - it's just pasting them together. Repeated column names get modified. What happens if the dataframes aren't of equal length?

```{r, error =TRUE}
data_4_rows <- tibble( names = letters[1:4], values = 1:4)
bind_cols(band_members, data_4_rows)
```

## Quiz

1. Examine the `table1` and `table4a` datasets. Combine  `table4a` to `table1` to create two new columns. Ensure the columns make sense and retain data integrity.
