# Tidying data

## About this chapter

1. Questions:
- How do I go from non-tidy to tidy data structures?
2. Objectives:
- Manipulating dataframes with the `tidyr` verbs
3. Keypoints:
- The `tidyr` package contains functions that affect the layout and structure of dataframes.

## tidyr

tidyr is a tool for manipulating layout of datasets. As part of the tidyverse it is loaded when you use `library(tidyverse)` but can be loaded on its own with `library(tidyr)`. tidyr has two main functions - `spread()` and `gather()`.


### Sample tidy datasets

Let's look at five sample tables that show the same data in different ways, only one of which counts as tidy.

```{r, echo=FALSE, warning = FALSE, message=FALSE}
library(tidyverse)
```



```{r}
table1
table2
table3
table4a
table4b
```

The tidy data is in `table1`. 

  * `table2` is not tidy because not every variable has its own column. The `count` column has values for `cases` and `population` and these are two different things. The `type` column _could_ be a variable on its own, but as used here its a way to mix up the count variable unneccesarily.
  * `table3` is not tidy because `rate` is currently a function of two variables - literally its printed as `cases/population`. The column `rate` should contain the result of `cases/population` and if we wanted to retain the `case` and `population` information it should be in its own column, like in `table1`
  * `table4a` and `table4b` aren't tidy, because the data are split over two tables and in each table the values of the `year` variable are split over multiple columns.
  
Let's work with each of these non-tidy datasets in turn to get them tidy.

## pivot_longer()

The most common problem is that in `table4a`, where the values of a variable are split over multiple columns. 

```{r}
table4a
```

To tidy this, we can use `pivot_longer()`, which increases the number of rows and decreases the number of columns. It needs three bits of information:

  1. The column(s) to keep un-pivoted - these are columns that are likely already tidy. All other columns will be pivoted
  2. The name of a new column in which to put the old column names
  3. The name of a new column in which to put the old values
  
The function call looks like this:

```{r}
table4a %>% 
  pivot_longer(-country, names_to = "year", values_to = "cases")
```

Note how we use the `-country` syntax to mean 'pivot everything but `country`'. The `names_to` argument tells `pivot_longer()` where to put the names, and the `values_to` argument specifies where the numbers should go.

Note too how the columns we pivoted (`1999` and `2000`) have been dropped from the table completely. This little table is now tidy. 

We can do the same with `table4b` but this one has the value `population`

```{r}
table4b %>% 
  pivot_longer(-country, names_to = "year", values_to = "population")
```

To combine these together we can use `left_join()`.
```{r}
t4a <- table4a %>% 
  pivot_longer(-country, names_to = "year", values_to = "cases")
t4b <- table4b %>% 
  pivot_longer(-country, names_to = "year", values_to = "population")

left_join(t4a, t4b)
```

Note we don't need to specify the `by = ` argument, since the two tables have column names in common - `left_join()` works that out and does the join automatically.


## pivot_wider()

The inverse function to `pivot_longer()` is `pivot_wider()` which increases column number and decreases row count. This function needs two pieces of information

  1. The column from which to get the new row names
  2. The columns from which to get the values
  
This is useful for dealing with the `table2` case.

```{r}
table2 %>% 
  pivot_wider(names_from = type, values_from = count)
```

## separate()

The `seperate()` function turns one column into many by splitting the value whenever a particular character appears. Remember `table3`

```{r}
table3
```

We can separate that `rate` column into two - `cases` and `population`

```{r}
table3 %>%
  separate(rate, into = c("cases", "population"))
```

By default `separate()` splits things on any non-numeric character. But we can be explicit with the `sep` argument.

```{r}
table3 %>%
  separate(rate, into = c("cases", "population"), sep = "/")
```

This works just as well, but is useful if the computer makes a bad guess.

Note the column types of `cases` and `population`, they're down as `chr`. By default the type of the parent column is retained, but you can make `separate()` guess what type the new column is with the `convert` argument.

```{r}
table3 %>%
  separate(rate, into = c("cases", "population"), sep = "/", convert = TRUE)
```
And now, we're back to tidy data.



## unite()

The `unite()` function is the inverse of `separate()`, and combines multiple columns into a single one.

To demonstrate `unite()` we can make a new table, `table5` by using the `separate()` function on the `year` column in `table3`. Passing `sep` a number, tells it just to split that many characters into the string  

Here's `table5`
```{r}
table5 <- table3 %>%
  separate(year, into = c("century", "year"), sep = 2, convert = TRUE)
table5
```

We can now re- `unite()` `table5`. The arguments for this function are just the name of the new column, and the columns to join

```{r}
table5 %>%
  unite(new, century, year)

```

Here the default is to use an underscore `_` to join the values, but we can be explicit and use nothing with the `sep` argument

```{r}
table5 %>%
  unite(new, century, year, sep="")

```

## Quiz

Tidying data is hard! And it needs you to know your data quite well, which naturally takes time. Rather than quiz questions here, a worked example will give good benefit, so let's try one of those. Do the Case Study on page 163 of R for Data Science. If you don't have the print edition it is available online here [Case Study for Tidy Data](http://r4ds.had.co.nz/tidy-data.html#case-study).


